The dynamics of risk perception in a Mediterranean agroecosystem

### 2020-02-19 manuscript/manuscript.Rmd
reorder sections

---
title: The dynamics of risk perception in a Mediterranean agroecosystem
subtitle: 
titlerunning: The dynamics of risk perception in a Mediterranean agroecosystem
authorrunning: Gauthier
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Nicolas Gauthier 1
  address: Laboratory of Tree-Ring Research & School of Geography and Development, University of Arizona
  email: ngauthier@email.arizona.edu

keywords:
- key
- dictionary
- word

#PACS: 
#- PAC1
#- superPAC

MSC:
- MSC code 1
- MSC code 2

@@

The secret to the success of earliest farmers on the mediterrenan littoral was in their manipulation of the local ecosystem to manage environmental risks. Early farmers constructed a complex agroecosystem to manage risk.
They needed to maintain stable crop yields in face of uncertain drought risks. Yield stability is very important for traditional agriculturlal commmunities not connected to larger bulk food networks (Abbo et al 2010)
neolithic crop package -- weiss zohary 2011
cite abbo et al 2010? for more stuff about ladraces
Wheat/barley systems
example of system that combines low yield, low wirks crop with high yield high risk crop, an effective context paut et al 2019
These strategies work when risks are knonw and constant, but may be more susceptible to periods of changing climateic risks
so understanding how people of the past adapted to change is important to us today

@@

The basic decision-making problem facing a farmer seeking to diversify their crops can be thought of as a game against nature.
We can represent this simplified decision context as a "game" in a game theorietic context (Luce raiffa 1989, milnor 1952, Agrawal1968, cassidy1971, gould 1963). The "game" in this context is the farmer's decision of which crops to plant and in what proportions, given uncertianty in the future "state-of-nature", and nature varies between several possible states such as dry and wet years (Table 1). In reality continuous, but people often intuitively solve an easier problem when faced with a complex real-world situation (cite behavioral econ or simon). The exact values in the payoff matrix here are less important than the relative payoffs in each quadrant. We can think of this game as being the culturally-inherited object. Our question is how do individuals make decisions here? So in this case the cultural norm would be to mix wheat and barley, but the learning is in the exact ratio.

@@

How should an early farmer make a decision in light ot this uncertainty?
The basic logic of crop diversitication is Portfolio of crops ala modern portfolio theory: Blank2001
Sometimes best to think of nature as a sentient opponent out to get you, and play strategically based on that assumption. Gould 1963,  Beckenkamp 2008
quick overview of criteria without risk, then say that a better option is to estimate the probabilities of the different state of nature
Agrawal1968 talks about the strategies in the context of agriculture being a game against nature
here I can slip in the cool thing about acting like the weather is out to get you being a good idea in states of complete uncertainty
transition -- the problem becomes easier if you have at least some idea of nature's moves, because then you cna work to get the highest yeilds given the risk of drought.
Decision strategies as choosing different points -- plot of decisions

@@

If an individual can learn about their environment, such as the risk of drought in any given year, then they can behave more rationally by trying to maximize their subjective expected utility. If the probability distributin of nature's moves is known, the farmer can choose the crop mix that simply maximizes the expected crop yields given the empirical frequency distirubtion of nature's moves. In the language of game theory, this strategy is known as a "game of ficticious play" against nature. This is an effective risk-managing strategy, but like all similar strategies that adapt to a specific pattern of variability, it is vulnerable to changes in the pattern of variability (Janssen eta al 2007)
This strategy works well when the environment is stationary -- nature plays from a fixed probability distirubtion, but is vulnerable to environmental nonstationarity. That is, when the mean or higher order moments of the rainfall distribution shift, playing a game of fictitious play can backfire because you're too stuck in your ways while the world changes around you. Thus it is important to understand the dynamics of risk perception -- how subjective risks, and thus subjective expected utility -- rise and fall in uncertain environments.

# Holocene drought risks in the eastern Mediterranean {#sec:2}

Estimates of drought risk in the eastern Mediterranean during the the past 10,000 years were derived from outputs of the TraCE-21k paleoclimate simulation available on the National Center for Atmospheric Research. TraCE-21K is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model (GCM) to recreate the transient response of the global climate system to changes in the Earth's orbit and greenhouse gasses over the past 22,000 years from the Last Glacial Maximum to the present [@He2011]. It generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of climate forcings (e.g. orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six hourly timescale, and model outputs are archived at a monthly resolution.

TraCE-21k simulation outputs for the past 10,000 years were extracted from the grid cell overlaying the northern Levant and southeastern Turkey. This location was selected due to its proximity to archaeological sites with comparative evidence of crop diversification strategies. This region includes some of the earliest evidence of crop domestication, including sites like Jericho and Can Hasan.

@@

@@

```{r, fig.height = 2, fig.width = 6, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period. The dashed line indicates the level of risk beyond which one would plant barley over wheat to maximize subjective expected yields, given the payoffs defined in Table 1.'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  filter(year != min(year)) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, size = 1) +
  geom_col()  +
  scale_fill_distiller(palette = 'PuOr') +
  scale_color_distiller(palette = 'PuOr')+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

# The dynamics of risk perception {#sec:3}

## Bayesian brains
We can efficiently simulate these population-level risk perceptions in a population of agents with Bayesian brains (Kahvalti et al 2019 double check). Bayesian learning allows for a minimal representation of the environemnt, so that each agent's beliefs can be encoded in two parameters -- defining the Beta distribution over their prior beliefs -- rather than the record of all observed years. Not only does this approach have a clear computational efficiency, but for the same reason this reflects the processes actually going on in the heads of decision makers.
Our brains do not record every bit of perceived informaiton in memory, rather it stores a "compact encoding" of that information which it uses for future decisions. Even if individuals aren't -- consciously or unconsciously --- making these calculations -- the basic algorithmic problem faced by the brain, and solutions it has evolved, reflect the same basic sets of constraints (Sanborn and Chater 2016).

```{r, fig.width = 5, fig.height = 3, fig.cap = "Development of an individual's perceived drought risk with time, assuming a fixed drought risk of 0.5. Beliefs are represented as Beta probabiity distributions, and the increased certainty with age reflects the varying effective sample size of the Beta prior."}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

### Bayesian learning from bernoulli observations
We can treat this problem elegantly using bayesian statistics using online bayesian learning from bernoulli objersbations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. bernoulli distribution) with paramter theta reprenting the risk of drought. Uncertainty in the value of theta can be represented with a beta distribution, constrained between 0 and 1 (Figure XX). We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief. What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely. As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of experience.

@@

### The problem of nonstationarity
That is, what if the risk of a drought, theta, is not a constant but rather varies unpredictably in time?
The statistical problem shifts then to one of change detection (o'reilly 2013, gallistel et al 2014).
In other words do people simply adapt to changes -- as in the fixed bayesian learner and moving average approaches -- or do they detect them?
Strict year-to-year updating can result in overly conservative risk perceptions, suggesting people are slow to change their beliefs when in reality they often quickly update them (Bissiri 2010).
Indeed, the "forgetful bayesian brain" model seems to perform better, with a dynamic rather than fixed bayesian belief model (Zhang and Yu 2013).

@@

So what are the consequences of these cognitive processes for early farmers in the Mediterranean.

```{r}
agents <- tibble(id = 1:150, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta),
                 #age = sample.int(80, size = 150, replace = TRUE),
                 crop = 'barley')
```

```{r}
learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

Simulate a time series with varying drought frequency. This is a stepwise nonstationary process. Setup an experiment exploring 6 scenarios.

@@

```{r}
make_droughts <- function() c(runif(25) < 0.2, runif(25) < 0.4, runif(25) < 0.2)

set.seed(11111)
droughts <- make_droughts()
```

```{r}
run_sim <- function(droughts) {
  accumulate(droughts, ~learn(.x, .y), .init = agents) %>% 
    bind_rows(.id = 'time') %>%
    mutate(time = as.numeric(time)) %>%
    filter(time > 1)
}

@@

```{r, fig.height = 8, fig.width = 7}

a <- ggplot(sim1) +
  geom_rect(aes(xmin = 25, xmax = 50, ymin = -Inf, ymax = Inf), fill = 'lightgray', alpha = .1) +
  geom_point(aes(time, risk), alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic() +
  labs(x = 'years', y = 'Perceived drought risk')

@@

# Discussion and Conclusions

@@

Examining the long-term dynamics of crop yields among all decision-making strategies and psychological profiles reveals variability in yields due to different decision making preferences is far greater variability due to differences in risk perception (Figure 7). Risk-minimizing strategies sacrifice productivity for predictability, consistently lowering the mean and variance of crop yields over time. Risk-minimizing strategies are more sensitive to risk perception than yield-maximizing strategies; varying the memory decay rate parameter $\lambda$ had no effect on the optimal crop allocation for yield maximizers.

>>>>

---
title: The dynamics of risk perception in a Mediterranean agroecosystem
subtitle: 
titlerunning: The dynamics of risk perception in a Mediterranean agroecosystem
authorrunning: Gauthier

authors: 
- name: Nicolas Gauthier 1
  address: Laboratory of Tree-Ring Research & School of Geography and Development, University of Arizona
  email: ngauthier@arizona.edu

keywords:
- drought risk
- belief-based agents
- crop diversification
- rapid climate change

@@

The secret to the success of earliest farmers on the mediterrenan littoral was in their manipulation of the local ecosystem to manage environmental risks. Early farmers constructed a complex agroecosystem to manage risk.
They needed to maintain stable crop yields in face of uncertain drought risks. Yield stability is very important for traditional agriculturlal commmunities not connected to larger bulk food networks (Abbo et al 2010)
neolithic crop package -- weiss zohary 2011
cite abbo et al 2010? for more stuff about ladraces
Wheat/barley systems
example of system that combines low yield, low wirks crop with high yield high risk crop, an effective context paut et al 2019
These strategies work when risks are knonw and constant, but may be more susceptible to periods of changing climatic risks
so understanding how people of the past adapted to change is important to us today

@@

The basic decision-making problem facing a farmer seeking to diversify their crops can be thought of as a game against nature.
We can represent this simplified decision context as a "game" in a game theorietic context (Luce raiffa 1989, milnor 1952, Agrawal1968, cassidy1971, gould 1963). The "game" in this context is the farmer's decision of which crops to plant and in what proportions, given uncertainty in the future "state-of-nature", and nature varies between several possible states such as dry and wet years (Table 1). In reality continuous, but people often intuitively solve an easier problem when faced with a complex real-world situation (cite behavioral econ or simon). The exact values in the payoff matrix here are less important than the relative payoffs in each quadrant. We can think of this game as being the culturally-inherited object. Our question is how do individuals make decisions here? So in this case the cultural norm would be to mix wheat and barley, but the learning is in the exact ratio.

@@

How should an early farmer make a decision in light of this uncertainty?
The basic logic of crop diversitication is Portfolio of crops ala modern portfolio theory: Blank2001
Sometimes best to think of nature as a sentient opponent out to get you, and play strategically based on that assumption. Gould 1963,  Beckenkamp 2008
quick overview of criteria without risk, then say that a better option is to estimate the probabilities of the different state of nature
Agrawal1968 talks about the strategies in the context of agriculture being a game against nature
here I can slip in the cool thing about acting like the weather is out to get you being a good idea in states of complete uncertainty
transition -- the problem becomes easier if you have at least some idea of nature's moves, because then you cna work to get the highest yeilds given the risk of drought.
Decision strategies as choosing different points -- plot of decisions

@@

If an individual can learn about their environment, such as the risk of drought in any given year, then they can behave more rationally by trying to maximize their subjective expected utility. If the probability distribution of nature's moves is known, the farmer can choose the crop mix that simply maximizes the expected crop yields given the empirical frequency distirubtion of nature's moves. In the language of game theory, this strategy is known as a "game of ficticious play" against nature. This is an effective risk-managing strategy, but like all similar strategies that adapt to a specific pattern of variability, it is vulnerable to changes in the pattern of variability (Janssen eta al 2007)

This strategy works well when the environment is stationary -- nature plays from a fixed probability distirubtion, but is vulnerable to environmental nonstationarity. That is, when the mean or higher order moments of the rainfall distribution shift, playing a game of fictitious play can backfire because as agents get too stuck in their ways while the world changes around them. Thus it is important to understand the dynamics of risk perception -- how subjective risks, and thus subjective expected utility -- rise and fall in uncertain environments.

# Modeling perceived risks with Bayesian agents {#sec:3}

We can efficiently simulate these population-level risk perceptions in a population of agents with Bayesian brains (Kahvalti et al 2019 ???double check). Bayesian learning allows for a minimal representation of the environment, so that each agent's beliefs can be encoded in two parameters -- defining the Beta distribution over their prior beliefs -- rather than the record of all observed years. Not only does this approach have a clear computational efficiency, but for the same reason this reflects the processes actually going on in the heads of decision makers.

Our brains do not record every bit of perceived information in memory, rather it stores a "compact encoding" of that information which it uses for future decisions. Even if individuals aren't -- consciously or unconsciously --- making these calculations -- the basic algorithmic problem faced by the brain, and solutions it has evolved, reflect the same basic sets of constraints (Sanborn and Chater 2016).

We can treat this problem elegantly using Bayesian statistics using online Bayesian learning from bernoulli observations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. bernoulli distribution) with parameter theta reprenting the risk of drought. Uncertainty in the value of theta can be represented with a beta distribution, constrained between 0 and 1. We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief \@ref(beta-prior). What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely. As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of experience.

```{r beta-prior, fig.width = 5, fig.height = 3, fig.cap = "Development of an individual's perceived drought risk with time, assuming a fixed drought risk of 0.5. Beliefs are represented as Beta probabiity distributions, and the increased certainty with age reflects the varying effective sample size of the Beta prior."}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

# Drought risks in the eastern Mediterranean {#sec:2}

Estimates of drought risk in the eastern Mediterranean during the the past 10,000 years were derived from outputs of the TraCE-21k paleoclimate simulation available on the National Center for Atmospheric Research. TraCE-21K is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model (GCM) to recreate the transient response of the global climate system to changes in the Earth's orbit and greenhouse gasses over the past 22,000 years from the Last Glacial Maximum to the present [@He2011]. It generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of external climate drivers (e.g. orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six hourly timescale, and model outputs are archived at a monthly resolution.

TraCE-21k simulation outputs for the past 10,000 years were extracted from the grid cell covering the northern Levant and southeastern Turkey. This location was selected due to its proximity to archaeological sites with comparative evidence of crop diversification strategies. This region includes some of the earliest evidence of crop domestication, including sites like Jericho and Can Hasan.

@@

Drought risks ranged from 1 in every 10 to 5 in 10 over the Holocene.

@@

```{r, fig.height = 2, fig.width = 6, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period. The dashed line indicates the level of risk beyond which one would plant barley over wheat to maximize subjective expected yields, given the payoffs defined in Table 1.'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  filter(year != min(year)) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, size = 1) +
  geom_col(width = 38)  +
  scale_fill_distiller(palette = 'PuOr') +
  scale_color_distiller(palette = 'PuOr')+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

@@

# Bayesian learning and the dynamics of risk perception in the Neolithic {#sec:4}

@@

So what are the consequences of these cognitive processes for early farmers in the Mediterranean?

Simulate a time series with varying drought frequency, based off the empirical drought risks derived from the paleoclimate simulation.

@@

That is, what if the risk of a drought, theta, is not a constant but rather varies unpredictably in time?
The statistical problem shifts then to one of change detection (o'reilly 2013, gallistel et al 2014).
In other words do people simply adapt to changes -- as in the fixed bayesian learner and moving average approaches -- or do they detect them?
Strict year-to-year updating can result in overly conservative risk perceptions, suggesting people are slow to change their beliefs when in reality they often quickly update them (Bissiri 2010).
Indeed, the "forgetful bayesian brain" model seems to perform better, with a dynamic rather than fixed bayesian belief model (Zhang and Yu 2013).

That is, the above figure assumes a noise free environment with a single learner. Next we can look at a population of learners in a noisy environment

Examining the long-term dynamics of crop yields among all decision-making strategies and psychological profiles reveals variability in yields due to different decision making preferences is far greater than variability due to differences in risk perception (Figure 7). Risk-minimizing strategies sacrifice productivity for predictability, consistently lowering the mean and variance of crop yields over time. Risk-minimizing strategies are more sensitive to risk perception than yield-maximizing strategies; varying the memory decay rate parameter $\lambda$ had no effect on the optimal crop allocation for yield maximizers. 
^^ change above to talk about individual uncertainty in drought risk is often greater than the actual variability in drought risk, given the limited life experience an individual draws from

```{r}
agents <- tibble(id = 1:150, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta),
                 #age = sample.int(80, size = 150, replace = TRUE),
                 crop = 'barley')

learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

make_droughts <- function() c(runif(25) < 0.2, runif(25) < 0.4, runif(25) < 0.2)

set.seed(11111)
droughts <- make_droughts()

run_sim <- function(droughts) {
  accumulate(droughts, ~learn(.x, .y), .init = agents) %>% 
    bind_rows(.id = 'time') %>%
    mutate(time = as.numeric(time)) %>%
    filter(time > 1)
}

@@

```{r, fig.height = 8, fig.width = 7}
a <- ggplot(sim1) +
  geom_rect(aes(xmin = 25, xmax = 50, ymin = -Inf, ymax = Inf), fill = 'lightgray', alpha = .1) +
  geom_point(aes(time, risk), alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic() +
  labs(x = 'years', y = 'Perceived drought risk')

@@

# Conclusion

@@
